---
title: "0. Getting started in R"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
---

# Introduction
In this pre-course exercise, you will:

- Create an R Project to use for this course
- Install all packages required for the course exercises
- Refresh your memory about some general operations in R


# Getting help
- Google it! Looking for help online is an integral part of modern coding. Any professional programmer who claims not to do this regularly is lying to you.
- All R functions should have documentation that you can access by typing `help(x)` or `?x`, where `x` is a function name.
- The [quick R website](https://www.statmethods.net/) has some excellent and brief tutorials on the most important topics.
- Hadley Wickham's book 'R for Data Science' provides a very gentle guide from the very basics to running your own analysis. You can read it for free at [here](http://r4ds.had.co.nz/).

# Getting started
### Install and load R
* The R language can be downloaded from [R project website](https://www.r-project.org/).
* It is highly recommended to write and perform analyses in [RStudio](https://www.rstudio.com/).
* RStudio allows you to create **projects** that keep all your work on a project together in a single folder. This helps enormously to keep your work organised and ideas together. To create a project for Ecological Methods:
    + Click `File` -> `New project`
    + In the window that opens click `New directory` then `New project`
    + Enter a name for the project and the folder that will contain it, for example `SDM-course`, and the directory where you want that folder to be located.

### The RStudio environment
* In a fresh RStudio session, the first thing to do is **open a new R script** by clicking `File` -> `New` -> `R script`, or through the the button with the green cross in the upper-left-hand corner.
* It is best to write your script in the *editor* window, and run sections of code using `Crtl`+`Return`. That way you will have a record of the code you have run so far.
* As you run code, it will appear in the *console* window
* As you progress, *objects* that you save (e.g., variables and tables) are listed in the *environment* window. When you close R, you will be asked if you want to save your workspace. This is not recommended; it is better to save the script that documents the analysis.
* The *plot* window is primarily for displaying graphics, and viewing *help* pages.
* RStudio can do **tab-completion** of function and object names, which is really handy. Start typing a name, for example the first few letters of the awkwardly named `getAllSuperClasses`, hit tab, and R will give you a list of suggestions.

### Basic concepts
* **Functions** tell R to perform an action, and are always followed by round brackets. For example:
```{r, eval=F}
c(1,2,3) # concatenate the numbers 1, 2 and 3
mean(c(1,2,3)) # take the average of 1, 2 and 3
```
* Functions take various **arguments** within the round brackets, that are separated by commas.
* We assign a name to data by using `<-` so we can use it later. For example:
```{r, eval=F}
x <- c(1,2,3)
mean(x)
```
* Square brackets are used to **subset** data. See the section on working with data for more...
* Sometimes we want to write **comments** for ourselves. The `#` symbol indicates that everything that follows on the same line is only a comment, and R should not try to evaluate it.
```{r, eval=F}
# First concatenate the values
x <- c(1,2,3)
mean(x) # Then take the mean
```


### Additional packages
* R has an enormous database of addition packages stored on the Comprehensive R Archive Network (CRAN).
* Install a package using `install.packages`. For example, we will need the following package to import data from Excel files:
```{r, eval=FALSE}
install.packages('readxl')
```

* Once installed, you can load the package using `library`:
```{r}
library(readxl)
```


# Working with data
### Working directory

* If you use an RStudio project, the working directory is automatically set to the project directory (this is the easy way to go!).
* When you import or export files, you specify their location relative to the working directory.
* You can find out what is your **current working directory** with:
```{r, eval=FALSE}
getwd()
```
* You can **change your working directory** by specifying a new path using something like:
```{r, eval=FALSE}
setwd('~/Documents/SDM-course')
```


